\stdpopsim: a community toolkit for population genetic simulation

### 2019-12-16 popSimManu.tex

\documentclass[12pt,halfline,a4paper]{ouparticle}
\usepackage{amsmath,amssymb}
\usepackage{natbib}
\usepackage{graphicx,rotating}
\usepackage{dsfont}
\usepackage{xspace}

@@

\section*{Introduction}
While population genetics has always used statistical methods to make inferences from data,
the degree of sophistication of the questions, models, data, and computational approaches
used have all increased over the past decade. Currently there exist myriad computational methods
that can infer the histories of populations, the distribution of fitness effects,
recombination rates, the extent of positive selection in genome sequence data
(\cite{li2011inference,schiffels2014inferring,terhorst2017robust,eyre2009estimating,kim2017inference,
chan2012genome,lin2013fast,Adrion662247,alachiotis2012omegaplus,degiorgio2016sweepfinder2,
kern2018diplos,sugden2018localization}).
While en masse these methods have increased our understanding of the
impacts of genetic and evolutionary processes, very little has been done to systematically
benchmark the quality of inferences gleaned from computational population genetics.
As large databases of population genetic variation begin to be used to inform public health procedures,
the accuracy and quality of these inferences is becoming even more important.
This is in large part because ``ground-truth'' in population genetics comes from simulation,
not from empirical observations, and as a field there has been no agreement as to what should
constitute community standards or best practices. The general modus operandi to date has been to
introduce methods which are validated by a set of bespoke simulations produced by individual
groups. Beyond a lack of efficiency from duplicated effort, such a situation leads
to decreases in reproducibility and transparency across the entire field.
Here we describe \stdpopsim, a community-driven catalog
of standardized simulations across a variety of common study organisms
that will empower the field and usher in the next era of genetics data analysis.

The degree to which modeling assumptions and different summaries of the
data can affect the ultimate results has been challenging to assess. Yet, there
have been several clear examples of different methods yielding fundamentally
different conclusions. For example, \MSMC methods applied to human genomes have
% JK is this right? > 100 years doesn't sound very ancient to me.
suggested large ancient ($>100$ years ago) ancestral population sizes and
bottlenecks that have not been detected by SFS-based methods (see
\citet{beichman2017comparison}). These distinct models differ in how they fit various summaries of the
genetic variation data, suggesting that modeling choices can greatly affect the
performance of the inference. Furthermore, it is reasonable
to assume generally that specific methods will perform better than others under certain scenarios.
However, because these conditions are typically unknown, empirical researchers
lack principled guidelines for deciding which statistical method is best suited
to accurately answer their particular question. The need for empirical
guidance will only increase as researchers continue to generate more genomic
variation data from non-model taxa and wish to make inferences regarding
demography and selection.

@@

\section*{Results}
\paragraph{The \stdpopsim library}
The central contribution from PopSim currently is \stdpopsim, a community
maintained library of empirical genome data and population genetics simulation
models, providing easy access to realistic simulations. Figure \ref{fig:cartoon} shows a graphical
representation of the structure of \stdpopsim. The package centers
on a catalog of species (Fig. \ref{fig:cartoon}A), initially consisting of humans, \emph{Drosophila},
\emph{Arabidopsis}, \emph{E. coli}, and \emph{Pongo}. A species definition consists
of two key elements.  Firstly, the library defines
some basic information about its genome, with information about chromosome
lengths, average mutation rates and so on. We also provide access to detailed
empirical information such as recombination maps, which model empirically
observed heterogeneity along chromosomes. As such maps are often large and hard
to find, we provide a mechanism to automatically download (and cache locally) a
set of known maps (Fig. \ref{fig:cartoon}B). For example, in humans we support the HapMapII
\cite{international2007second}) and
deCODE (\cite{kong2010fine}) genetic maps. The second key element of a species description
within \stdpopsim is a set of carefully curated population genetic model
descriptions from the literature, which are the basis for accurate simulations
of specific historical scenarios.

Given the genome data and simulation model descriptions defined within the
library, it is then straightforward to run accurate, standardized simulations
across a range of organisms. \stdpopsim has a Python API and a user-friendly
command line interface (Fig. \ref{fig:cartoon}C), allowing users with minimal experience direct access to
state-of-the-art simulations. Simulations are output in the “tree sequence”
format (\cite{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}), which
contains complete genealogical information about the simulated samples, is
extremely compact, and can be processed very efficiently using the tskit library
(\cite{kelleher2016efficient,kelleher2018efficient}; Fig. \ref{fig:cartoon}D). Currently,
\stdpopsim uses the  \texttt{msprime} coalescent simulator (\cite{kelleher2016efficient})
as the simulation engine. We plan to include other simulation
engines to support scenarios that cannot be modelled under the coalescent framework,
and have already implemented \texttt{SLiM} (\cite{haller2019slim}) as
an alternative back-end.

@@

\subsection*{\texttt{stdpopsim} Catalog}
The current contents of the \texttt{stdpopsim} catalog are shown in Table X. At the time of
writing the PopSim Consortium has implemented previously published demographic models for
humans, \emph{Drosophila}, and \emph{Arabidopsis}. These include both
simpler, single population histories (e.g., in \emph{D. melanogaster} \cite{sheehan2016deep}),
and much more complex models which include population splitting, migration, and archaic
admixture (e.g., in humans \cite{ragsdale2019models}).

@@

We start by comparing popular methods for estimating
population size histories ($N(t)$) of single populations and subsequently
show simple examples of multi-population inference.
To evaluate and compare the performance of inference methods we developed
\texttt{snakemake} (\cite{koster2012snakemake}) workflows that allow reproducibility
in comparing inference methods which are available from \url{https://github.com/popgensims/analysis}
. The workflows allow efficient computing in multicore environments or on
cluster environments.

For single-population population size histories, we compared \MSMC, \smcpp, and
\stairwayplot (\cite{schiffels2014inferring,terhorst2017robust,liu2015exploring})
 on simulated genomes sampled from a single population,
in a number of the \stdpopsim models described above. The pipeline was generalized to
run $R$ replicates with $C$ chromosomes for $N$ samples, for a total of $R \times C$
simulations for each demographic model. After simulation, the pipeline prepares
input files for each of the respective inference methods by grouping all
chromosomes, for each sample, into a single file. This step results in an
input file for each simulation replicates $R$, pertaining to each of the
respective inference methods and derived from the same simulated tree sequences.
Each of the inference programs are then run in parallel, followed by plotting of
$N_e$ estimates from each program.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Ragsdale.png}
\caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
size over time ($N(t)$) inferred using 4 different methods, \smcpp, \texttt{stairwayplot},
\MSMC with $n=2$ and $n=8$. Data were generated by simulating
replicate human genomes under the \cite{ragsdale2019models} model and using the genetic map
inferred in \cite{international2007second}. From top to bottom we show estimates for each
of the three populations in the model YRI, CEU, and CHB. In shades of blue we show the estimated
$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
for the rate of coalescence in the demographic model.}
\label{fig:n_t_ragsdale}
\end{center}
\end{figure}

In Figure \ref{fig:n_t_ragsdale} we present results from our $N(t)$ analysis pipeline
run on whole genome simulations under the \cite{ragsdale2019models} model using the
human genetic map estimated from \cite{international2007second}. On rows
we show inferred $N(t)$ for the three extant populations in the model.
On columns we show comparisons among the methods (note that we compare two differing
sample sizes for \MSMC). The solid black lines represent the true effective population
size as inferred from the inverse coalescent rate defined by the model.
Blue lines represent estimates from each of three replicate whole genome simulations.
While there is variation among methods and individual population estimates,
generally spoken the methods are accurate for this model of human history.

\stdpopsim allows us to readily compare these estimates to those drawn from a different
model of human history. In Figure \ref{fig:n_t_gutenkunst} we show estimates of
$N(t)$ from simulations using the same physical and genetic maps, but a different demographic
history-- that inferred by \cite{gutenkunst2009inferring}. Again we see that each
of the methods is capturing relevant parts of the population history, although the
degree of noise at any given time interval varies. In comparing inferences between the
models it is interesting to note that $N(t)$ estimates for the CHB and CEU
simulated populations are generally better across methods than estimates from the YRI
sample.

@@

As most population genetic methods development has been focused on human-scale
data, it is of consequence to ask how such methods might perform in non-human
genomes. Figure \ref{fig:n_t_sheehan} shows parameter estimates from a demographic
model estimated from an African sample of \emph{Drosophila melanogaster} (\cite{sheehan2016deep}).
As with humans, we use \stdpopsim to simulate replicate \emph{Drosophila} genomes under
an empirically derived genetic map, and then try to infer back the parameters of model.
Accuracy is mixed among methods in this setting, and generally worse than what we
observe under simulations of the human genome.

@@

\subsection*{Workflow for two population analysis}
Our goal for the two population analysis pipeline was to explore two population inference
on some of the multi-population demographic models implemented in \stdpopsim.
The models we examined include the human Gutenkunst model (\cite{gutenkunst2009inferring}) as
well as the \emph{Drosophila} Li and Stephan model (\cite{li2006inferring}). The human
model is a three population model (Africa, Europe, and Asia) including post-divergence
migration and exponential growth (Figure \ref{fig:IM_popn_human}C), whereas the
\emph{Drosophila} model is a two population model (Africa and Europe) with no post-divergence
migration and constant population sizes (Figure \ref{fig:two_popn_fly}).

To conduct inference on these models, we implemented three commonly-used approaches:
\dadi (\cite{gutenkunst2009inferring}), \fastsimcoal2 (\cite{excoffier2013robust}),
and \smcpp (\cite{terhorst2017robust}). Our aim was not to exhaustively
test the performance of these methods, but rather to explore their behavior
in two different inference scenarios and demonstrate the potential for this resource to
eventually be used for more in-depth benchmarking. As a result, these methods were used
generally with default settings and we did not attempt to optimize their performance or fit
parameter-rich demographic models. As with the single population inference pipeline,
we developed a \texttt{snakemake} (\cite{koster2012snakemake}) workflow to manage
our inference pipeline and maintain reproducibility.

@@

To perform inference, we began by running simulations under each model, running
three replicates to inspect for inter-replicate variability. For humans, we ran
whole genome simulations (chromosomes 1-22) under the Gutenkunst model (\cite{gutenkunst2009inferring}),
using the HapMap II recombination map (\cite{international2007second}) and taking 20
samples each from the Europe and Africa populations. For \emph{Drosophila},
we simulated 3 Mb of chromosome 2R under the Li and Stephan model (\cite{li2006inferring}),
using the map inferred by \cite{comeron2012many} and taking 50 samples from each of
the Africa and Europe populations. We tried to run whole genome simulations under
this model but after XXXX amount of time... (JRA fill in details here). For the generic
IM simulations, we used the human genome (chromosomes 1-22) and HapMap II recombination
map (\cite{international2007second}) and took 20 samples per population.

>>>>

\documentclass[12pt,halfline,a4paper]{ouparticle}
\usepackage{amsmath,amssymb}
\usepackage[round]{natbib}
\usepackage{graphicx,rotating}
\usepackage{dsfont}
\usepackage{xspace}

@@

\section*{Introduction}
While population genetics has always used statistical methods to make inferences from data,
the degree of sophistication of the questions, models, data, and computational approaches
used have all increased over the past decade. Currently there exist myriad computational methods
that can infer the histories of populations, the distribution of fitness effects,
recombination rates, the extent of positive selection in genome sequence data
\citep{li2011inference,schiffels2014inferring,terhorst2017robust,eyre2009estimating,kim2017inference,
chan2012genome,lin2013fast,Adrion662247,alachiotis2012omegaplus,degiorgio2016sweepfinder2,
kern2018diplos,sugden2018localization}.
While en masse these methods have increased our understanding of the
impacts of genetic and evolutionary processes, very little has been done to systematically
benchmark the quality of inferences gleaned from computational population genetics.
As large databases of population genetic variation begin to be used to inform public health procedures,
the accuracy and quality of these inferences is becoming even more important.
This is in large part because ``ground-truth'' in population genetics comes from simulation,
not from empirical observations, and as a field there has been no agreement as to what should
constitute community standards or best practices. The general modus operandi to date has been to
introduce methods which are validated by a set of bespoke simulations produced by individual
groups. Beyond a lack of efficiency from duplicated effort, such a situation leads
to decreases in reproducibility and transparency across the entire field.
Here we describe \stdpopsim, a community-driven catalog
of standardized simulations across a variety of common study organisms
that will empower the field and usher in the next era of genetics data analysis.

The degree to which modeling assumptions and different summaries of the
data can affect the ultimate results has been challenging to assess. Yet, there
have been several clear examples of different methods yielding fundamentally
different conclusions. For example, \MSMC methods applied to human genomes have
% JK is this right? > 100 years doesn't sound very ancient to me.
suggested large ancient ($>100$ years ago) ancestral population sizes and
bottlenecks that have not been detected by SFS-based methods
\citep[see][]{beichman2017comparison}.
These distinct models differ in how they fit various summaries of the
genetic variation data, suggesting that modeling choices can greatly affect the
performance of the inference. Furthermore, it is reasonable
to assume generally that specific methods will perform better than others under certain scenarios.
However, because these conditions are typically unknown, empirical researchers
lack principled guidelines for deciding which statistical method is best suited
to accurately answer their particular question. The need for empirical
guidance will only increase as researchers continue to generate more genomic
variation data from non-model taxa and wish to make inferences regarding
demography and selection.

@@

\section*{Results}
\paragraph{The \stdpopsim library}
The central contribution from PopSim currently is \stdpopsim, a community
maintained library of empirical genome data and population genetics simulation
models, providing easy access to realistic simulations. Figure \ref{fig:cartoon} shows a graphical
representation of the structure of \stdpopsim. The package centers
on a catalog of species (Fig. \ref{fig:cartoon}A), initially consisting of humans, \emph{Drosophila},
\emph{Arabidopsis}, \emph{E. coli}, and \emph{Pongo}. A species definition consists
of two key elements.  Firstly, the library defines
some basic information about its genome, with information about chromosome
lengths, average mutation rates and so on. We also provide access to detailed
empirical information such as recombination maps, which model empirically
observed heterogeneity along chromosomes. As such maps are often large and hard
to find, we provide a mechanism to automatically download (and cache locally) a
set of known maps (Fig. \ref{fig:cartoon}B). For example, in humans we support the
HapMapII~\citep{international2007second} and
deCODE~\citep{kong2010fine} genetic maps. The second key element of a species description
within \stdpopsim is a set of carefully curated population genetic model
descriptions from the literature, which are the basis for accurate simulations
of specific historical scenarios.

Given the genome data and simulation model descriptions defined within the
library, it is then straightforward to run accurate, standardized simulations
across a range of organisms. \stdpopsim has a Python API and a user-friendly
command line interface (Fig. \ref{fig:cartoon}C), allowing users with minimal experience direct access to
state-of-the-art simulations. Simulations are output in the “tree sequence”
format \citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}, which
contains complete genealogical information about the simulated samples, is
extremely compact, and can be processed very efficiently using the tskit library
\citep{kelleher2016efficient,kelleher2018efficient}. Currently,
\stdpopsim uses the  \texttt{msprime} coalescent simulator \citep{kelleher2016efficient}
as the simulation engine. We plan to include other simulation
engines to support scenarios that cannot be modelled under the coalescent framework,
and have already implemented \texttt{SLiM} \citep{haller2019slim} as
an alternative back-end.

@@

\subsection*{\texttt{stdpopsim} Catalog}
The current contents of the \texttt{stdpopsim} catalog are shown in Table X. At the time of
writing the PopSim Consortium has implemented previously published demographic models for
humans, \emph{Drosophila}, and \emph{Arabidopsis}. These range from
simple, single population histories \cite[e.g.,][]{sheehan2016deep},
to complex models which include population splitting, migration, and archaic
admixture \cite[e.g.,][]{ragsdale2019models}.

@@

We start by comparing popular methods for estimating
population size histories ($N(t)$) of single populations and subsequently
show simple examples of multi-population inference.
To evaluate and compare the performance of inference methods we developed
\texttt{snakemake} \citep{koster2012snakemake} workflows that allow reproducibility
in comparing inference methods which are available from \url{https://github.com/popgensims/analysis}
. The workflows allow efficient computing in multicore environments or on
cluster environments.

For single-population population size histories, we compared
\MSMC~\citep{schiffels2014inferring}, \smcpp~\citep{terhorst2017robust}, and
\stairwayplot~\citep{liu2015exploring}
 on simulated genomes sampled from a single population,
in a number of the \stdpopsim models described above. The pipeline was generalized to
run $R$ replicates with $C$ chromosomes for $N$ samples, for a total of $R \times C$
simulations for each demographic model. After simulation, the pipeline prepares
input files for each of the respective inference methods by grouping all
chromosomes, for each sample, into a single file. This step results in an
input file for each simulation replicates $R$, pertaining to each of the
respective inference methods and derived from the same simulated tree sequences.
Each of the inference programs are then run in parallel, followed by plotting of
$N_e$ estimates from each program.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Ragsdale.png}
\caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
size over time ($N(t)$) inferred using 4 different methods, \smcpp, \texttt{stairwayplot},
\MSMC with $n=2$ and $n=8$. Data were generated by simulating
replicate human genomes under the \cite{ragsdale2019models} model and using the
HapMapII genetic map~\citep{international2007second}. From top to bottom we show estimates for each
of the three populations in the model YRI, CEU, and CHB. In shades of blue we show the estimated
$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
for the rate of coalescence in the demographic model.}
\label{fig:n_t_ragsdale}
\end{center}
\end{figure}

% JK here --- I think it would really help if we described what the models
% were, briefly, rather than saying who inferred them.
In Figure \ref{fig:n_t_ragsdale} we present results from our $N(t)$ analysis pipeline
run on whole genome simulations under a sophisticated demographic model
and empirical genetic map. On rows
we show inferred $N(t)$ for the three extant populations in the model.
On columns we show comparisons among the methods (note that we compare two differing
sample sizes for \MSMC). The solid black lines represent the true effective population
size as inferred from the inverse coalescent rate defined by the model.
Blue lines represent estimates from each of three replicate whole genome simulations.
While there is variation among methods and individual population estimates,
generally spoken the methods are accurate for this model of human history.

\stdpopsim allows us to readily compare these estimates to those drawn from a different
model of human history. In Figure \ref{fig:n_t_gutenkunst} we show estimates of
$N(t)$ from simulations using the same physical and genetic maps, but a different demographic
history -- that inferred by \cite{gutenkunst2009inferring}. Again we see that each
of the methods is capturing relevant parts of the population history, although the
degree of noise at any given time interval varies. In comparing inferences between the
models it is interesting to note that $N(t)$ estimates for the CHB and CEU
simulated populations are generally better across methods than estimates from the YRI
sample.

@@

As most population genetic methods development has been focused on human-scale
data, it is of consequence to ask how such methods might perform in non-human
genomes. Figure \ref{fig:n_t_sheehan} shows parameter estimates from a demographic
model estimated from an African sample of \emph{Drosophila melanogaster} \citep{sheehan2016deep}.
As with humans, we use \stdpopsim to simulate replicate \emph{Drosophila} genomes under
an empirically derived genetic map, and then try to infer back the parameters of model.
Accuracy is mixed among methods in this setting, and generally worse than what we
observe under simulations of the human genome.

@@

\subsection*{Workflow for two population analysis}
Our goal for the two population analysis pipeline was to explore two population inference
on some of the multi-population demographic models implemented in \stdpopsim.
The models we examined include the human Gutenkunst model \citep{gutenkunst2009inferring} as
well as the \emph{Drosophila} Li and Stephan model \citep{li2006inferring}. The human
model is a three population model (Africa, Europe, and Asia) including post-divergence
migration and exponential growth (Figure \ref{fig:IM_popn_human}C), whereas the
\emph{Drosophila} model is a two population model (Africa and Europe) with no post-divergence
migration and constant population sizes (Figure \ref{fig:two_popn_fly}).

To conduct inference on these models, we implemented three commonly-used approaches:
\dadi~\citep{gutenkunst2009inferring}, \fastsimcoal2~\citep{excoffier2013robust},
and \smcpp~\citep{terhorst2017robust}. Our aim was not to exhaustively
test the performance of these methods, but rather to explore their behavior
in two different inference scenarios and demonstrate the potential for this resource to
eventually be used for more in-depth benchmarking. As a result, these methods were used
generally with default settings and we did not attempt to optimize their performance or fit
parameter-rich demographic models. As with the single population inference pipeline,
we developed a \texttt{snakemake}~\citep{koster2012snakemake} workflow to manage
our inference pipeline and maintain reproducibility.

@@

To perform inference, we began by running simulations under each model, running
three replicates to inspect for inter-replicate variability. For humans, we ran
whole genome simulations (chromosomes 1-22) under the Gutenkunst model \citep{gutenkunst2009inferring},
using the HapMap II recombination map \citep{international2007second} and taking 20
samples each from the Europe and Africa populations. For \emph{Drosophila},
we simulated 3 Mb of chromosome 2R under the Li and Stephan model \citep{li2006inferring},
using the map inferred by \cite{comeron2012many} and taking 50 samples from each of
the Africa and Europe populations. We tried to run whole genome simulations under
this model but after XXXX amount of time... (JRA fill in details here). For the generic
IM simulations, we used the human genome (chromosomes 1-22) and HapMap II recombination
map \citep{international2007second} and took 20 samples per population.
